---
title: "A3: Incarceration"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(maps)
library(mapdata)
```

```{r}
# Load datasets
prison_pop <- read_csv("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv")
```

```{r}
colnames(prison_pop)
```

### Introduction

The trends.vera.org data tool is utilized for gathering county-level prison and 
jail population statistics to gain a clearer understanding of the current 
incarceration landscape in the U.S. The focus of the analysis is on examining trends
in incarceration within the U.S. I have selected the Dataset [Dataset about prison population per county/state](https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-pop.csv?raw=true). 
This dataset offers extensive information on the prison population, including 
variables such as the total prison population, prison population by race (Black,
White, Latinx, etc.), and by gender. These variables are essential for comprehending 
the disparities and trends in incarceration among different demographic groups. 
By analyzing this data, we can investigate how incarceration rates differ by race 
and gender, and how these trends have changed over time, thereby providing insights 
into systemic issues and highlighting areas that require policy intervention. The 
inclusion of detailed demographic data makes this dataset an invaluable resource 
for identifying patterns of inequality and guiding potential reforms in the criminal
justice system.



### Summary Information

The analysis of prison populations across counties reveals that the average total 
prison population is approximately 361.36 individuals. The highest recorded white 
prison population in a single county is 9945, while the lowest is 0, indicating 
significant disparities. Additionally, the average male prison population is around 
459.57 individuals. Over the past five years, there has been an average decrease 
of 1.53 in the total prison population, suggesting a slight overall decline in 
incarceration rates. These findings underscore the variability in incarceration 
across different demographics and regions.



```{r}
# 1. Average total prison population across all counties
average_total_prison_pop <- prison_pop %>%
  summarise(avg_total_prison_pop = mean(total_prison_pop, na.rm = TRUE))

# 2. Highest white prison population
highest_white_prison_pop <- prison_pop %>%
  filter(white_prison_pop == max(white_prison_pop, na.rm = TRUE)) %>%
  select(white_prison_pop)

# 3.Lowest White prison population
lowest_white_prison_pop <- prison_pop %>%
  filter(white_prison_pop == min(white_prison_pop, na.rm = TRUE)) %>%
  select(white_prison_pop)

# 4. Average of change in the total prison population over the past 5 years
change_over_time <- prison_pop %>%
  group_by(year) %>%
  summarise(avg_prison_pop = mean(total_prison_pop, na.rm = TRUE)) %>%
  mutate(change = avg_prison_pop - lag(avg_prison_pop)) 

recent_change <- change_over_time %>%
  filter(year %in% (max(year) - 4):max(year))

# Calculate the average change over the past 10 years
average_recent_change <- recent_change %>%
  summarise(avg_change = mean(change, na.rm = TRUE))



# 5. Average male prison population
average_male_prison_pop <- prison_pop %>%
  summarise(avg_male_prison_pop = mean(male_prison_pop, na.rm = TRUE))

# Print the summary values
average_total_prison_pop
highest_white_prison_pop
lowest_white_prison_pop
average_recent_change
average_male_prison_pop
```




### The Dataset

Who collected the data?  

The data was gathered from multiple sources, including the Census of Jails, Census 
of State and Federal Adult Correctional Facilities, FBI Uniform Crime Reporting 
Program, National Corrections Reporting Program, Deaths in Custody Reporting 
Program, Annual Survey of Jails, Centers for Disease Control, U.S. Department of
Agriculture, and U.S. Census Bureau. This effort involved extensive collaboration
among various organizations.



How was the data collected or generated?  

The data was compiled through the collaboration of many organizations, which also 
supplemented it with additional information to fill in any gaps.



Why was the data collected?  

The data was collected to comprehensively understand incarceration trends at the 
county level in the U.S. This allows for a broad understanding of jail and prison 
populations, though it may lack specific and precise details.



How many observations (rows) are in your data?

There are 153,811 observations (rows).

How many features (columns) are in the data?  

There are 36 features (columns).

What, if any, ethical questions or questions of power do you need to consider 
when working with this data? 

We should take this data seriously, recognizing that while there are significant 
differences in incarceration rates between races, it is too simplistic to conclude 
that this indicates racism or a racial problem. The situation is much more complex. 
Additionally, a high incarceration rate is not a positive indicator, so we hope 
to see a decreasing trend, which would suggest improved safety in the U.S.



What are possible limitations or problems with this data?   (at least 200 words)

The dataset on prison populations per county/state from the Vera Institute of 
Justice, while comprehensive, has several limitations and potential issues. A 
major limitation is the presence of missing or incomplete data, as not all 
correctional facilities consistently report their data. This inconsistency can 
lead to gaps, affecting the reliability and accuracy of the analysis and causing 
significant variations in how different states and counties report their incarceration
statistics, introducing potential biases and inaccuracies. Additionally, the dataset 
is historical and may not fully reflect recent changes in incarceration policies 
or practices. As criminal justice reform efforts gain momentum, changes such as 
reducing mandatory minimum sentences or decriminalizing certain offenses may not 
be captured, potentially misleading analyses of the current state of incarceration. 
Another issue is the lack of qualitative data to provide deeper insights into the 
reasons behind trends. Quantitative data alone cannot fully explain the socio-economic 
and political factors contributing to incarceration rates. Factors such as changes 
in law enforcement practices, economic conditions, and social movements, which are 
not captured in the dataset, play crucial roles. Systemic biases within the criminal 
justice system, like over-policing in minority communities and discriminatory 
sentencing practices, can also influence the data, leading to an over-representation 
of certain racial groups, particularly Black and Latinx populations, potentially 
skewing the analysis and reinforcing negative stereotypes. Moreover, the dataset 
does not account for the lived experiences of incarcerated individuals or the broader 
impact of incarceration on communities, leading to an incomplete understanding of 
the issue and hindering the development of holistic and effective policy solutions. 
Finally, interpreting the data requires careful consideration. Statistical analyses 
can highlight disparities but do not always explain the underlying causes. Simplistic 
conclusions drawn from complex data can misinform policy decisions. It is essential 
to approach the data with a critical and nuanced perspective, acknowledging its 
limitations and considering the broader socio-economic and political contexts 
contributing to incarceration trends. In summary, while the dataset is valuable 
for understanding incarceration trends in the U.S., its limitations and potential
biases must be carefully managed to ensure accurate and ethical use.



### Trends Over Time Chart

``````{r}
trend_data_gender_race <- prison_pop %>%
  filter(year >= 1990) %>%
  group_by(year) %>%
  summarise(
    black_male_prison_pop = sum(black_male_prison_pop, na.rm = TRUE),
    black_female_prison_pop = sum(black_female_prison_pop, na.rm = TRUE),
    white_male_prison_pop = sum(white_male_prison_pop, na.rm = TRUE),
    white_female_prison_pop = sum(white_female_prison_pop, na.rm = TRUE),
    latinx_male_prison_pop = sum(latinx_male_prison_pop, na.rm = TRUE),
    latinx_female_prison_pop = sum(latinx_female_prison_pop, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(black_male_prison_pop, black_female_prison_pop, white_male_prison_pop, white_female_prison_pop, latinx_male_prison_pop, latinx_female_prison_pop), names_to = "group", values_to = "prison_pop")

trend_data_gender_race <- trend_data_gender_race %>%
  separate(group, into = c("race", "gender"), sep = "_", extra = "merge")

ggplot(trend_data_gender_race, aes(x = year, y = prison_pop, color = interaction(race, gender), group = interaction(race, gender))) +
  geom_line(size = 1) +
  labs(
    title = "Trends in U.S. Prison Populations by Race and Gender (1990 - Present)",
    x = "Year",
    y = "Prison Population",
    color = "Race and Gender"
  ) +
  theme_minimal()
```
This chart illustrates the prison population trends for male and female individuals
within Black, White, and Latinx groups from 1990 to the most recent year available, 
highlighting significant disparities among these groups. It provides a comprehensive 
view of how incarceration rates vary by both race and gender, aiming to inform users 
about the intersectional dynamics in the prison system. The chart reveals that the 
Black male prison population significantly increased from 1990 to around 2010 before 
plateauing and declining, reflecting the impact of the War on Drugs and subsequent
reforms. Similar trends are observed for Black females, White males, and White 
females, with increases until around 2010, followed by stabilization and declines.
The Latinx male and female prison populations show stable increases until around 
2005, then declines, potentially due to changes in immigration policies. This 
visualization offers insights into the intersectional disparities within the prison 
system and the impact of various policies and societal changes on these groups 
over time.





### Variable Comparison Chart

This chart illustrates the relationship between the total prison population and 
the female prison population, visually representing how the female prison population 
correlates with the overall prison population. The inclusion of this chart is 
crucial for analyzing gender dynamics within the prison system and identifying 
patterns that may inform policy decisions and reforms. The chart reveals a positive 
correlation between the total prison population and the female prison population, 
with the female population increasing as the total population increases. The linear 
regression line (in blue) indicates a strong linear relationship between the two 
variables, suggesting that changes in the total prison population are reflected
in changes in the female prison population. This visualization provides a clear 
picture of how the female prison population scales with the total prison population,
highlighting the proportional representation of female inmates and assisting in 
understanding the broader gender-related trends within the prison system.


```{r, warning = FALSE, message = FALSE}
prison_pop <- read_csv("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv")
chart2_data <- prison_pop %>%
  filter(!is.na(total_prison_pop) & !is.na(female_prison_pop))

ggplot(data = chart2_data, aes(x = total_prison_pop, y = female_prison_pop)) +
  geom_point(color = "grey") +
  geom_smooth(method=lm, color= "blue",  se=FALSE) +
  labs(title = paste("Relationship Between Prison Population and Female",
                     "Prison Population"),
       x = "Total Prison Population",
       y = "Female Prison Population")
```


### Map

This map shows the geographic distribution of the total prison population across 
different states in the United States, visually representing how prison populations 
are distributed geographically and providing insights into regional patterns and 
disparities. The inclusion of this map aims to offer a clear visual representation 
of these patterns, which is crucial for identifying areas with high incarceration 
rates and informing policy decisions to address these disparities. The map reveals 
significant regional concentrations of prison populations, with states in darker 
shades of blue indicating higher total prison populations. There is considerable 
variability across states, highlighting disparities influenced by state-specific 
policies, economic conditions, and other factors. This visualization helps identify 
regions with higher incarceration rates and contributes to a better understanding 
of the factors influencing these patterns.




```{r, warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(maps)

prison_pop <- read_csv("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv")

state_prison_pop <- prison_pop %>%
  group_by(state) %>%
  summarise(total_prison_pop = sum(total_prison_pop, na.rm = TRUE))

states_map <- map_data("state")

state_prison_pop$state <- tolower(state.name[match(state_prison_pop$state, state.abb)])

map_data_prison <- merge(states_map, state_prison_pop, by.x = "region", by.y = "state", all.x = TRUE)

ggplot(data = map_data_prison, aes(x = long, y = lat, group = group, fill = total_prison_pop)) +
  geom_polygon(color = "white") +
  scale_fill_continuous(name = "Total Prison Population", low = "lightblue", high = "darkblue") +
  labs(
    title = "Geographic Distribution of Prison Population by State",
    x = "",
    y = ""
  ) +
  coord_map() +
  theme_minimal()
```

